---
title: "ISL-chp2"
author: "KS"
date: '2018-11-06'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("ISLR")
library("gridExtra")
library("broom")
library("MASS")
```

### Question 8

Exercise on College data set.

```{r College}
college <- as.tibble(College)
names(college) <- tolower(names(college))
glimpse(college)
```

### Summary

You can also embed plots, for example:

```{r college}
summary(college)
```

### Plot 'outstate versus private' (boxplot)
Plot shows that outstate tuition is higher at private colleges.

```{r}
p1 <- ggplot(data = college, aes(x = private, y = outstate)) + 
    geom_boxplot()
```

### Create new factor variable 'elite' 
('Yes' if Top10perc > 50 else 'No')

```{r}
college <- college %>% 
    mutate(elite = as_factor(if_else(top10perc > 50, "Yes", "No")))
#select(college, top10perc, elite)
```

### Plot 'outstate versus elite' (boxplot)
Plot shows that outstate tuition is higher at elite institutions.

```{r}
p2 <- ggplot(data = college, aes(x = elite, y = outstate)) +
    geom_boxplot()
```

### Plot grid

```{r}
p1 <- ggplot(data = college, aes(x = f.undergrad)) +
    geom_histogram(bins = 50)
p2 <- ggplot(data = college, aes(x = expend, fill = private)) + 
    geom_histogram(bins = 50)
p3 <- ggplot(data = college, aes(x = accept/apps, fill = elite)) + 
    geom_histogram(bins = 50)
p4 <- ggplot(data = college, aes(x = room.board, fill = private)) + 
    geom_histogram(bins = 30)
grid.arrange(p1, p2, p3, p4, nrow = 2, ncol = 2)
```

### 

```{r}
ggplot(data = college, aes(x = private, y = accept/apps)) +
    geom_boxplot()
```

```{r}
ggplot(data = college, aes(x = room.board, y = outstate, colour = private)) +
    geom_point()
```

Graduation rate at elite universities is higher. 

```{r}
college %>% 
    filter(grad.rate <= 100) %>% 
    ggplot(aes(x = elite, y = grad.rate)) +
    geom_boxplot()
```

Graduation rate is higher at private colleges

```{r}
college %>%
    filter(grad.rate <= 100) %>% 
    ggplot(aes(x = private, y = grad.rate)) +
    geom_boxplot()
```

Slightly more students from top 10% of high school class go to private colleges

```{r}
ggplot(data = college, aes(x = private, y = top10perc)) +
    geom_boxplot()
```

Student - faculty ratio is lower at private colleges

```{r}
ggplot(data = college, aes(x = private, y = s.f.ratio)) +
    geom_boxplot()
```

Student - faculty ratio is lower at private colleges

```{r}
ggplot(data = college, aes(x = elite, y = s.f.ratio)) +
    geom_boxplot()
```


### Question 9

Exercise on Auto data set.

```{r}
auto <- as_tibble(Auto)
glimpse(auto)
summary(auto)
```

name: qualitative. cylinders, origin: may also be treated as ordered qualitative variables.

use sapply to apply the function to the data by columns.

```{r}
sapply(auto[,1:8], range, na.rm = TRUE)
```

```{r}
round(sapply(auto[,1:8], mean, na.rm = TRUE), 2)
```

```{r}
round(sapply(auto[,1:8], sd, na.rm = TRUE), 2)
```

Using map_dbl instead of sapply

```{r}
map(auto[,1:8], range)
round(map_dbl(auto[,1:8], mean), 3)
round(map_dbl(auto[,1:8], sd), 3)
```

```{r}
auto_reduced <- auto[-(10:85),]
sapply(auto_reduced[,1:8], range)
#map(auto_reduced[,1:8], range)
round(map_dbl(auto_reduced[,1:8], mean), 3)
round(map_dbl(auto_reduced[,1:8], sd), 3)
```

```{r}
png(filename = "figures/pairs_auto.png")
pairs(auto[,1:8])
dev.off()
```

```{r}
auto$cylinders <- as.factor(auto$cylinders)
auto$origin <- as.factor(auto$origin)
```

### Question 10

```{r}
boston <- as.tibble(Boston)
glimpse(boston)
```

```{r}
dim(boston)
pairs(boston[,1:7])
```

```{r}
ggplot(data = boston, mapping = aes(x = age, y = nox)) +
    geom_point()
ggplot(data = boston, mapping = aes(x = medv, y = lstat)) +
    geom_point()
ggplot(data = boston, mapping = aes(x = black, y = crim)) +
    geom_point()
ggplot(data = boston, mapping = aes(x = crim)) + 
    geom_histogram()
ggplot(data = boston, mapping = aes(x = tax)) + 
    geom_histogram()
ggplot(data = boston, mapping = aes(x = ptratio)) + 
    geom_histogram()
```

```{r}
sum(boston$chas == 1)
median(boston$ptratio)
boston[which(boston$medv == min(boston$medv)), ]
summary(boston)
```

```{r}
sum(boston$rm > 7)
sum(boston$rm > 8)
boston[(boston$rm > 8),]
```

